---
title: "Class 11 EC"
format: gfm
editor: visual
---

# Section 4: Population Scale Analysis

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
table <- read.table("rs8067378_ENSG00000172057.6.txt")
```

```{r}
graph <- boxplot(table$exp[table[,"geno"]=="A/A"], 
                 table$exp[table[,"geno"]=="A/G"], 
                 table$exp[table[,"geno"]=="G/G"], 
                 xlab = "Genotypes", ylab = "Expression Values")

graph
```

```{r}
# The third row in the stats column of boxplot results are the medians of each genotype. 
graph$stats[3,]
```

The median expression level for A\|A genotype is 31.2, for A\|G is 25.1, and for G\|G is 20.1.

```{r}
# The n column tells you the observations of each boxplot.
graph$n
```

The sample size for A\|A is 108, for A\|G is 233, and for G\|G is 121.

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?
>
> Looking at the boxplot from Q13 we can see that the expression of the gene ORMDL3 is lower in people with the G\|G genotype compared to the A\|A genotype because the G\|G group has a lower median and a distribution of expression values that are near or below the median of A\|A.

```{r}
boxplot(table$exp[table[,"geno"]=="A/A"], 
        table$exp[table[,"geno"]=="A/G"], 
        table$exp[table[,"geno"]=="G/G"], 
        xlab = "Genotypes", ylab = "Expression Values")

```
