---
title: "Class 10: Halloween Mini-Project"
author: "Sarah Tareen"
format: gfm
editor: visual
---

# 1. Importing candy data

Let's get the data:

```{r}
candy = read.csv('candy-data.csv', row.names=1)
head(candy)
```

> **Q1**. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

There are 85 candy types in the dataset.

> **Q2**. How many fruity candy types are in the dataset?

```{r}
table(candy$fruity)
```

There are 38 fruity candy types.

# 2. What is your favorite candy?

> **Q3**. What is your favorite candy in the dataset and what is it's `winpercent` value?

```{r}
candy["Skittles original", ]$winpercent
```

The winpercent for Skittles original is 63.%

> **Q4**. What is the `winpercent` value for "Kit Kat"?

```{r}
candy["Kit Kat", ]$winpercent
```

The winpercent for Kit Kat is 77%.

> **Q5**. What is the `winpercent` value for "Tootsie Roll Snack Bars"?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

The winpercent for Tootsie Roll Snack Bars is 50%.

There is a helpful package for getting a brief overview of the dataset.

```{r}
#install.packages("skimr")
library("skimr")
skim(candy)
```

> -   **Q6**. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes the columns n_missing and complete_rate are on the scale of only ones and zeroes while the rest of the columns are between 0 and 1.

> -   **Q7**. What do you think a zero and one represent for the `candy$chocolate` column?

I believe a one means that the candy has chocolate in it and a zero means it doesn't.

Now we can do exploratory data analysis.

> **Q8**. Plot a histogram of `winpercent` values

```{r}
hist(candy$winpercent)
```

> **Q9**. Is the distribution of `winpercent` values symmetrical?

No the distribution is not symmetrical because it is skewed more towards the left and the median is below 50%.

> **Q10**. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

The mean is slightly above 50%.

> **Q11**. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# We want to find the mean of the winpercents for a specific type of candy. We can make a subset of the winpercent column by turning the ones and zeroes into trues and falses. 

chocolate_candy = candy$winpercent[as.logical(candy$chocolate)]
mean(chocolate_candy)

fruity_candy = candy$winpercent[as.logical(candy$fruity)]
mean(fruity_candy)
```

Chocolate candy is ranked higher on average than fruity candy.

> **Q12**. Is this difference statistically significant?

```{r}
# We are 95% confident that the difference between these two values is there. At minimum, we are confident they are 11% different and at the maximum they are 22% different. 
t.test(chocolate_candy, fruity_candy)
```

Yes this difference is statistically significant because the t test has a very low p-value of 2.871e-08.

We can also see this difference qualitatively using a boxplot.

```{r}
boxplot(chocolate_candy, fruity_candy)
```

# 3. Overall Candy Rankings

> **Q13**. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```

The 5 least liked candy types are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.

> **Q14**. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent, decreasing = TRUE),], n=5)
```

The 5 all time favorite candy types are Reese's Peanut Cups, Reese's Miniature Cups, Twix, Kit Kat, and Snickers.

> **Q15**. Make a first barplot of candy ranking based on `winpercent` values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> **Q16**. This is quite ugly, use the `reorder()` function to get the bars sorted by `winpercent`?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

Let's add color:)

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill=my_cols)
```

> **Q17**. What is the worst ranked chocolate candy?
>
> The worst ranked chocolate candy is Sixlets.
>
> \
> **Q18**. What is the best ranked fruity candy?
>
> The best ranked fruity candy is Starburst.

# 4. Taking a look at pricepercent

Let's compare candy rank and price.

```{r}
#add text labels that don't overlap
#install.packages("ggrepel")
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

We can see which candy is most and least expensive by ordering the columns.

> **Q19**. Which candy type is the highest ranked in terms of `winpercent` for the least money - i.e. offers the most bang for your buck?

Looking at the plot, we can see that Reese's Miniatures are one of the highest ranking candies (considering a winpercent of 80% is good) for the least money.

> **Q20**. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

The top 5 most expensive candy types are Nestle Smarties, Nik L Nip, Ring Pop, Hershey's Krackel, and Hershey's Milk Chocolate. The least popular one is Hershey's Milk Chocolate.

Now let's make a plot of pricepercent.

```{r}
ggplot(candy) + 
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_col(fill=my_cols)
```

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

# 5. Exploring the correlation structure

```{r}
#install.packages("corrplot")
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> **Q22**. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Two variables that are anti-correlated are chocolate and fruity.

> **Q23**. Similarly, what two variables are most positively correlated?

Chocolate and bar are the most positively correlated candy types.

# 6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Let's plot the first two components.

```{r}
attributes(pca)
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
p
```

```{r}
#install.packages('plotly')
library(plotly)

#ggplotly(p)
#commented out because interactive plots won't render
```

![](images/Screenshot%20(23).png)

```{r}
#Let's check the loadings of the PCA.
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> **Q24**. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
>
> The variables fruity, hard, and pluribus are picked up strongly by PC1 in the positive direction. This makes sense because I can think of multiple examples of candies that do have all or two of these attributes at the same time like Starburst and Skittles. So these variables are positively correlated.
