---
title: "Class 17 - Covid"
author: "Sarah Tareen"
format: gfm
editor: visual
---

# Getting Started

```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test (1).csv")
head(vax)
```

> **Q1.** What column details the total number of people fully vaccinated?
>
> persons_fully_vaccinated
>
> **Q2.** What column details the Zip code tabulation area?
>
> zip_code_tabulation_area
>
> **Q3.** What is the earliest date in this dataset?

```{r}
min(vax$as_of_date)
```

> **Q4.** What is the latest date in this dataset?

```{r}
vax[nrow(vax),1]
```

Let's take a look at the data.

```{r}
library(skimr)
skimr::skim_without_charts(vax)
```

> **Q5.** How many numeric columns are in this dataset?
>
> 14 numeric columns as we can see from the `skim` result.

```{r}
# check how the data frame is structured
str(vax)
```

```{r}
#check a specific column
class(vax$persons_fully_vaccinated)
```

> **Q6.** Note that there are "missing values" in the dataset. How many `NA` values there in the `persons_fully_vaccinated` column?

```{r}
sum(is.na(vax$persons_fully_vaccinated))
```

> **Q7.** What percent of `persons_fully_vaccinated` values are missing (to 2 significant figures)?

```{r}
(17711 / 220500)*100
```

> **Q8.** \[Optional\]: Why might this data be missing?
>
> Individual data points are not perfect and working with a huge amount of data has outliers which are not super important to the overall trend.

## Working with dates

```{r}
library(lubridate)
```

What is today's date?

```{r}
today()
```

```{r}
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

```{r}
today() - vax$as_of_date[1]
```

> **Q9.** How many days have passed since the last update of the dataset?

```{r}
today() - vax$as_of_date[nrow(vax)]
```

> **Q10.** How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
length(unique(vax$as_of_date))
```

# Working with ZIP codes

```{r}
library(zipcodeR)
```

```{r}
geocode_zip('92037')
```

We can find the distance between the centers of any two ZIP codes in miles.

```{r}
zip_distance('92037','92109')
```

We can find information about specific zip codes.

```{r}
reverse_zipcode(c('92037', "92109") )
# Pull data for all ZIP codes in the dataset
#zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )
```

# Focus on the San Diego area

There are two ways to focus on the San Diego area:

We can use base R functions.

```{r}
# Subset to San Diego county only areas
sd <- vax[ vax$"county" == "San Diego",]
```

Or we can use the `dplyr` package.

```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```

`dplyr` is useful when you want multiple filters...

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```

> **Q11.** How many distinct zip codes are listed for San Diego County?

```{r}
length(unique(sd$zip_code_tabulation_area))
```

107 unique zip codes for San Diego county.

> Q12. What San Diego County Zip code area has the largest population in this dataset?

```{r}
largest_zip_code <- sd[sd$tot_population == max(sd$tot_population, 
                                                na.rm = TRUE),]

unique(largest_zip_code$zip_code_tabulation_area)
```

92154 has the largest San Diego population.

> **Q13.** What is the overall average (with 2 decimal numbers) "Percent of Population Fully Vaccinated" value for all San Diego "County" as of "2023-05-23"?

```{r}
sd_recent <- filter(sd, as_of_date == "2023-05-23")
mean(sd_recent$percent_of_population_fully_vaccinated, na.rm = TRUE)
```

> **Q14.** Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of "2023-05-23"?

```{r}
#Base R
hist(sd_recent$percent_of_population_fully_vaccinated, 
     xlab = "Percent of population fully vaccinated", 
     main = "Vaccination Rates Across\nSan Diego County on May 23, 2023")

#ggplot2
library(ggplot2)
ggplot(sd_recent) + aes(x = percent_of_population_fully_vaccinated) + geom_histogram() + 
  labs(x= "Percent of population fully vaccinated", 
       y = "Count = Zip Codes", 
       title = "Vaccination Rates",
       subtitle = "Across San Diego County on May 23, 2023")
```

## Focus on UCSD/La Jolla

Let's filter to the UC San Diego in the 92037 ZIP code area and with an age 5+ population size of 36,144.

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

> **Q15**. Using **ggplot** make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
plot_92037 <- ggplot(ucsd) +
  aes(x=as_of_date,
      y=percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
plot_92037
```

## Comparing to similar sized areas

> **Q16**. Calculate the mean *"Percent of Population Fully Vaccinated"* for ZIP code areas with a population as large as 92037 (La Jolla) *as_of_date* "2023-05-23". Add this as a straight horizontal line to your plot from above with the `geom_hline()` function?

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2023-05-23")

head(vax.36)
```

```{r}
mean(vax.36$percent_of_population_fully_vaccinated)
```

```{r}
plot_92037 + geom_hline(yintercept=0.7226674, linetype="dashed", color="red")
```

> **Q17.** What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the *"Percent of Population Fully Vaccinated"* values for ZIP code areas with a population as large as 92037 (La Jolla) *as_of_date* "2023-05-23"?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

> **Q18.** Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) + aes(x=percent_of_population_fully_vaccinated) + geom_histogram() + 
  labs(x="Percent Vaccinated",
       title="Vaccination Rate Across ZIP codes with a population 
       as large as 92037 (La Jolla)",
       subtitle= "Shown for 2023-05-23")

```

> **Q19**. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2023-05-23") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```

```{r}
vax %>% filter(as_of_date == "2023-05-23") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```

The percentage of people fully vaccinated for the 92109 and 92040 ZIP code areas are below the average value of the ZIP code areas with a population as large as 92037 (La Jolla).

> **Q20.** Finally make a time course plot of vaccination progress for all areas in the full dataset with a `age5_plus_population > 36144`.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(x=as_of_date,
      y=percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="violetred2") +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination Rate Across California",
       subtitle= "Only areas with a population above 36k are shown.") +
  geom_hline(yintercept = 0.7225892, linetype= "dashed")

```
